---
title: 'R Exercises: Statistical Shape Analysis'
author: "Ian Dryden"
subtitle: Complex Data School, June 24-25, Seoul, Korea
output:
  pdf_document:
    toc: true
    toc_depth: '2'
  html_document:
    toc: true
    toc_depth: 2
    number_sections: false
---

```{r chunk 1, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE
)
```

# Notes

This document contains the R exercise code for the short course on Shape Analysis. 

Several exercises require the R packages `shapes` and `fdasrvf`.

# R Exercise 1: Bookstein shape co-ordinates

Compute and explore Bookstein shape coordinates for the mouse landmark data.

```{r chunk 2}
library("shapes")
data(mice)
u<-bookstein2d( mice$x )$bshpv
plotshapes( u , col=as.double(mice$group) , symbol=1:6)
legend("topright",c("control","large","small"),pch=1,col=c(1,2,3))
```

# R Exercise 1: Bookstein shape co-ordinates — part 2

Compute and explore Bookstein shape coordinates for the mouse landmark data.

```{r chunk 3}
ub <- t(u[3:6,1,]) ; vb <- t(u[3:6,2,])
pairs( cbind( centroid.size( mice$x), ub, vb) ,
  labels=c("S","u3","u4","u5","u6","v3","v4","v5","v6"), 
     col=as.double(mice$group) )
```

# R Exercise 2: Procrustes shape distance

Work through pairwise Procrustes alignment and compare the manual calculation with package functions.

```{r chunk 4}
centre <- function(X) scale(X, scale = FALSE)
csize  <- function(X) sqrt(sum(centre(X)^2))
preshape <- function(X) centre(X) / csize(X)
X <- mice$x[,,1]
Y <- mice$x[,,31]
X0 <- preshape(X)
Y0 <- preshape(Y)
sv <- svd(t(X0) %*% Y0)
R  <- sv$v %*% t(sv$u)
R
Y_aligned <- Y0 %*% R
dP <- sqrt(sum((X0 - Y_aligned)^2))
dP
print( sqrt( 2*(1-cos( riemdist( X,Y) ) )))   #check
print( riemdist( X,Y) )                           # similar
```

# R Exercise 3: Pairwise Procrustes Fit 

Work through pairwise Procrustes alignment and compare the manual calculation with package functions.

```{r chunk 5}
X <- mice$x[,,1]
Y <- mice$x[,,31]
outline <- c(1,6,2,3,4,5,1)
plotshapes( abind( X, Y)  , joinline=outline )

#match Y to X, without scaling
opa1 <- procOPA( X, Y, scale=FALSE)
opa1
plotshapes( abind( opa1$Ahat , opa1$Bhat) , joinline=outline )

#match Y to X, with scaling
opa2 <- procOPA( X, Y, scale=TRUE) 
opa2
plotshapes( abind( opa2$Ahat , opa2$Bhat) , joinline=outline )
```

# R Exercise 4: GPA with the shapes package

Use Generalized Procrustes Analysis to align configurations and estimate a mean shape.

```{r chunk 6}
data(mice)
# mice$x is k x m x n landmark array
gpa <- procGPA(mice$x, scale=TRUE)
aligned <- gpa$rotated.  # aligned configurations
mean_shape <- gpa$mshape   # mean shape

plotshapes(gpa$rotated)
lines( mean_shape[c(1,6,2,3,4,5,1),], col=2)

#Some DNA data 
data(dna.dat)
proc <- procGPA( dna.dat , scale=FALSE)
shapes3d( proc$rotated )
```

# R Exercise 5: Digit 3 data in R

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 7}
library(shapes)  #  load the `shapes' package (may need installing)
digit3.dat[,,1:2]           #  Will print out the coordinates of the digit3 data
plotshapes(digit3.dat)  # all 30 observations - what a mess!
plotshapes(digit3.dat,joinline=1:13)  # a bit less of a mess
```

# R Exercise 5: Digit 3 data in R — part 2

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 8}
out <- procGPA( digit3.dat , scale = TRUE) 
out$rotated[,,1:2]     # contains the Procrustes registered data
plotshapes ( out$rotated , joinline=c(1:13) )   # Procrustes registered data
out$mshape    # contains the Procrustes mean shape
lines( out$mshape, col=2 )  # adds the mean shape to the plot (in red)
plot( out$size , out$rho )  
```

# R Exercise 5: Digit 3 (continued)

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 9}
plot( out$size , out$rho)
text( out$size , out$rho , c(1:30) )
```

# R Exercise 5: Digit 3 (continued)

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 10}
plotshapes( digit3.dat , digit3.dat[,,1] , joinline= 1:13)
out2 <- procGPA( digit3.dat[-1,,]  , scale = TRUE)   #Procrustes w/o the outlier
plot( out2$size , out2$rho)
text( out2$size , out2$rho , c(2:30) )
```

# R Exercise 6: Digit 3 outlier

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 11}
out<-procGPA( digit3.dat ) 
shapepca(out,type="r",mag=1,joinline=1:13)  # plot the effects of the first 3 PCs 
shapepca(out,type="v",mag=1)
shapepca(out,type="g",mag=1)
```

# R Exercise 6: Digit 3 outlier — part 2

Inspect, register, and analyse the digit 3 landmark data, including the effect of an outlier.

```{r chunk 12}
plot(out$rho)
out2<-procGPA( digit3.dat[-1,,] )      # without outlier 
shapepca(out2,type="r",mag=1)  # plot the effects of the first 3 PCs 
shapepca(out2,type="v",mag=1)
shapepca(out2,type="g",mag=1)
```

# R Exercise 7: T2 mice - three groups

Analyse the three mouse groups after GPA and compare tangent-coordinate choices.

```{r chunk 13}
data(mice)
gpa <- procGPA(mice$x)
shapepca(gpa,type="r",mag=3)
shapepca(gpa,type="v",mag=3)
shapepca(gpa,type="g",mag=3)
plot(gpa$scores,col=as.factor(mice$group),pch=as.character(mice$group))
gpa2 <- procGPA(mice$x,tangentcoords="expomap")
plot(gpa2$scores,col=as.factor(mice$group),pch=as.character(mice$group))
```

# R Exercise 8: 3D Monkey skulls

Read and analyse the 3D monkey skull data; this requires the external file `monkey.csv`.

```{r chunk 14}
tem <- read.csv("monkey.csv") 
tem<-as.matrix(tem)
dim(tem) #Note the first column has a landmark label (number) in it. 
tem1 <- tem[,2:58]
monk <- array(tem1, c(7, 3, 19) )
```

# R Exercise 8: 3D Monkey skulls — part 2

Read and analyse the 3D monkey skull data; this requires the external file `monkey.csv`.

```{r chunk 15}
library(shapes)
ans <- procGPA( monk)
ans$percent
plot(ans$rho)
shapepca( ans )
```

# R Exercise 8 (continued)

Run this chunk after loading any required data and packages.

```{r chunk 16}
ans2 <- pnss3d( monk , n.pc=10, sphere.type="small")
ans2$percent
out <- plot3darcs( ans2, pcno=1, c=3)
```

# R Exercise 8 (continued)

Run this chunk after loading any required data and packages.

```{r chunk 17}
ans3 <- pnss3d( monk[,,-19] , n.pc=10, sphere.type="small")
out <- plot3darcs( ans3, pcno=1, c=3)
```

# R Exercise 9: DNA

Apply PCA and principal nested spheres-style analysis to the DNA dataset.

```{r chunk 18}
data(dna.dat)
proc <- procGPA( dna.dat )
proc$percent
shapepca(proc)

ans <- pnss3d( dna.dat , n.pc=10, sphere.type="BIC")
ans$percent
temp <- plot3darcs(ans, pcno=1, c=2)
```

# R Exercise 10: Peptides

Analyse peptide configurations; this requires the external file `peptide.Rda`.

```{r chunk 19}
load("peptide.Rda")
proc <- procGPA( z1 )
proc$percent
shapepca(proc)
plot(proc$scores[,1],proc$scores[,2])
```

# R Exercise 10: Peptides — part 2

Analyse peptide configurations; this requires the external file `peptide.Rda`.

```{r chunk 20}
ans <- pnss3d( z1 , n.pc=10, sphere.type="BIC")
ans$percent
plot(ans$PNS$scores[,1],ans$PNS$scores[,2])
lines(ans$PNS$scores[,1],ans$PNS$scores[,2],col=1:100)
tem <- plot3darcs(ans, pcno=1, c=1, type="pnss")
#compare to PCA
tem2 <-plot3darcs(ans, pcno=1, c=1, type="pca")
```

# R Exercise 11: permutation test

Compare mean shapes between mouse groups using permutation and resampling tests.

```{r chunk 21}
mlarge <- mice$x[,,mice$group=="l"]
msmall <- mice$x[,,mice$group=="s"]
testmeanshapes( mlarge, msmall )
resampletest( mlarge, msmall )
mcontrol <- mice$x[,,mice$group=="c"]
testmeanshapes( mlarge, mcontrol )
resampletest( mlarge, mcontrol )
```

# R Exercise 12: Fisher-Rao growth velocity data

Analyse growth velocity curves using Fisher–Rao registration and functional PCA tools.

```{r chunk 22}
library("fdasrvf")
data(growth_vel)
n<-93
palette(rainbow(n+10))
plot(growth_vel$time,growth_vel$f[,1],ylim=c(-2,30),type="n",ylab="f")
for (i in 1:n){
lines(growth_vel$time,growth_vel$f[,i],col=n+10-i)
}
out <- time_warping( growth_vel$f, growth_vel$time)
plot(growth_vel$time,growth_vel$f[,1],ylim=c(-2,30),type="n",ylab="f")
for (i in 1:n){
lines(growth_vel$time,out$fn[,i],col=n+10-i)
}
```

# R Exercise 12: Fisher-Rao growth velocity (continued)

Analyse growth velocity curves using Fisher–Rao registration and functional PCA tools.

```{r chunk 23}
hfpca <- horizFPCA(out, no = 3)
vfpca <- vertFPCA(out, no = 3)
jfpca <- jointFPCA(out, no = 3)
```


# R Exercise 13: Statistical shape analysis of six groups DNA data

This R code reads in the data from 12 groups of DNA molecular dynamics simulations 
(undamaged XGX and damaged XFX) and then calculates a few simple summaries from the 
thinned AFA group. 

```{r chunk 24}
library(shapes)
afa<-read.in("./DNA/AFA_P.x",22,3) ; afc<-read.in("./DNA/AFC_P.x",22,3) 
afg<-read.in("./DNA/AFG_P.x",22,3) ; aga<-read.in("./DNA/AGA_P.x",22,3)  
agc<-read.in("./DNA/AGC_P.x",22,3) ; agg<-read.in("./DNA/AGG_P.x",22,3)
tfa<-read.in("./DNA/TFA_P.x",22,3) ; tfc<-read.in("./DNA/TFC_P.x",22,3) 
tft<-read.in("./DNA/TFT_P.x",22,3) ; tga<-read.in("./DNA/TGA_P.x",22,3) 
tgc<-read.in("./DNA/TGC_P.x",22,3) ; tgt<-read.in("./DNA/TGT_P.x",22,3)

ans<-procGPA(afa,scale=FALSE) #investigate AFA group
shapes3d(ans$rotated)
plot(ans$rho) #shape distances `rho' from the mean shape
plot(ans$size) #size
ts.plot(ans$size) # time series plot
acf(ans$size)
pacf(ans$size)
```
